---
title: "Serum analysis"
date: "`r Sys.Date()`"
graphics: yes
output:
  html_document:
    theme: cerulean
    code_folding: hide
  pdf_document: default
  word_document: default
header-includes: \usepackage{xcolor}
vignette: "%\\VignetteEngine{knitr::rmarkdown} %\\VignetteEncoding{UTF-8}\n"
---

# Introduction

This is the data analysis pipeline accompanying the article Dewald et al. 2022. We start by loading libraries, simulated data (with some dependencies) and setting parameters. The simulated data is based on the distributions of the observed data from the study. Variable dependencies are fictional and not equivalent to the observed relations.

# Installation

The R software can be freely acquired from https://www.r-project.org/. The installation of R and the packages can take several minutes due to a large number of dependencies. Dependencies are automatically installed.

```{r,eval=FALSE}
install.packages("BiocManager")

BiocManager::install(c("bnlearn","parallel","abind","mnem","devtools","lmtest"))

devtools::install_github("MartinFXP/RcppFun")
```

```{r,echo=TRUE,fig.height=6}
start <- Sys.time()
library(bnlearn)
library(RcppFun)
library(parallel)
cl <- makePSOCKcluster(2)
invisible(clusterEvalQ(cl,"library(mnem); library(bnlearn)"))
graph2adj <- mnem:::graph2adj
clusterExport(cl,c("as.graphNEL","graph2adj","hc"))
library(abind)
library(mnem)
scaled <- 1
n <- 100
m <- 0.5
redo <- 0
docpdag <- 1
df <- readRDS("serum_data_simulated.rds")
for (i in 1:ncol(df)) {
  if (class(df[,i])=="character") {
    df[,i] <- factor(df[,i])
  }
}
print(str(df))
```

# Step-wise forward regression with likelihood-ratio test

In each step we create a new model for each un-selected variable and compare to the currently best model by testing the significance of the model fit. The best variable is then added to the model. All others are again tested until none is significant anymore or all are included in the best model.

```{r,echo=TRUE,fig.height=6}
par(mfrow=c(1,2))
responses <- c("SpikeLoad","ID50Wuh01","ID50BA5","ID50BQ11")
featureVals <- as.data.frame(matrix(NA,ncol(df),length(responses)))
rownames(featureVals) <- colnames(df)
colnames(featureVals) <- responses
counter <- 0
useVarsList <- list()
for (response in responses) {
  counter <- counter + 1
  dftmp <- df
  if (response=="SpikeLoad") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BA5","ID50BQ11"))]
  }
  if (response=="ID50Wuh01") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50BA5","ID50BQ11"))]
  }
  if (response=="ID50BA5") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BQ11"))]
  }
  if (response=="ID50BQ11") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BA5"))]
  }
  useVars <- response
  formula <- as.formula(paste0(response," ~ 1"))
  modBest <- lm(formula, data=dftmp)
  formula <- as.formula(paste0(response," ~ ."))
  fitinc <- NULL
  coltype <- NULL
  testVars <- colnames(dftmp)[colnames(dftmp)!=response]
  while(length(testVars)>0) {
    pvals <- numeric(length(colnames(dftmp)[which(colnames(dftmp) %in% testVars)]))
    for (i in testVars) {
      mod <- lm(formula, data=dftmp[,c(useVars,i)])
      pvals[which(i==colnames(dftmp)[which(colnames(dftmp) %in% testVars)])] <- (lmtest::lrtest(mod,modBest)[5])[2,]
    }
    fitinc <- c(fitinc,-log10(min(pvals)))
    names(fitinc)[length(fitinc)] <- colnames(dftmp)[which(colnames(dftmp) %in% testVars)][which.min(pvals)]
    if (min(pvals)<0.05) {
      useVars <- c(useVars,colnames(dftmp)[which(colnames(dftmp) %in% testVars)][which.min(pvals)])
      coltype <- c(coltype,1)
    } else {
      coltype <- c(coltype,2)
    }
    testVars <- testVars[which(!testVars %in% names(fitinc))]
    modBest <- lm(formula, data=dftmp[,useVars])
  }
  useVarsList[[response]] <- useVars
  par(mar=c(3,8,2,2))
  cols <- rep(rgb(1,0,0),sum(coltype==1))
  cols <- c(cols,rep(rgb(0.9,0.9,0.9),sum(coltype==2)))
  barplot(rev(fitinc),horiz=TRUE,las=1,col=rev(cols),
          xlab="-log10(p-value)",
          main=paste0(response,"\nadding most significant (<0.05) variable"))
  abline(v=-log10(0.05))
  featureVals[match(names(fitinc),rownames(featureVals)),counter] <- fitinc
}
```

# Show final models

In the final models we can scale the variable to make the coefficients comparable.

```{r,echo=TRUE}
models <- list()
count <- 1
counter <- 0
for (response in c("SpikeLoad","ID50Wuh01","ID50BA5","ID50BQ11")) {
  counter <- counter + 1
  formula <- as.formula(paste0(response," ~ ."))
  useVars <- useVarsList[[response]]
  dftmp <- df[,useVars]
  if (scaled) {
    dftmp[,unlist(lapply(dftmp,class))!="factor"] <- scale(dftmp[,unlist(lapply(dftmp,class))!="factor"])
  }
  mod <- lm(formula, data=dftmp)
  print(paste0(responses,c("","","","~Wuh01","","~Wuh01+BA5"))[counter])
  print(summary(mod))
  models[[counter]] <- mod
}
```

# Bayesian network analysis

We run n bootstraps and summarize them by counting the edges. The final network is plotted with a cut off for edge fraction by m.

```{r,echo=TRUE,fig.width=12,fig.height=4}
adj2dnf <- function(A) {
    dnf <- NULL
    for (i in seq_len(ncol(A))) {
        dnf <- c(dnf, rownames(A))
        for (j in seq_len(nrow(A))) {
            if (A[i, j] > 0) {
                dnf <- c(dnf, paste(colnames(A)[i], rownames(A)[j], sep = "="))
            }
            if (A[i, j] < 0) {
                dnf <- c(dnf, paste("!", colnames(A)[i], "=", rownames(A)[j],
                                    sep = ""))
            }
        }
    }
    dnf <- unique(dnf)
    return(dnf)
}
counter <- 0
for (response in c("SpikeLoad","ID50Wuh01","ID50BA5","ID50BQ11")) {
  counter <- counter + 1
  dftmp <- df
  if (response=="SpikeLoad") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BA5","ID50BQ11"))]
  }
  if (response=="ID50Wuh01") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50BA5","ID50BQ11"))]
  }
  if (response=="ID50BA5") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BQ11"))]
  }
  if (response=="ID50BQ11") {
    dftmp <- df[,which(!colnames(df) %in% c("ID50Wuh01","ID50BA5"))]
  }
  useVars <- c(response,useVarsList[[response]])
  bl <- NULL
  for (i in 1:ncol(dftmp)) {
    bl <- rbind(bl,c(response,colnames(dftmp)[i]))
  }
  colnames(bl) <- c("from","to")
  bl <- as.data.frame(bl)
  bn <- hc(dftmp,blacklist=bl,restart=100,max.iter=1)
  bnorg <- mnem:::graph2adj(as.graphNEL(bn))
  bnbs <- bnorg*0
  doBs <- function(i,bnbs,df,bl) {
    bnbs <- hc(df[sample(1:nrow(df),nrow(df),replace=TRUE),],blacklist=bl,
               restart=100,max.iter=1000)
    return(bnbs)
  }
  filename <- paste0(responses[counter],"_cpdag_",docpdag,".rds")
  if (file.exists(filename) & redo==0) {
    bnbs <- readRDS(filename)
  } else {
    bnbs <- clusterApply(cl,1:n,doBs,bnbs,dftmp,bl)
    saveRDS(bnbs,filename)
  }
  bnbs <- lapply(bnbs,function(x,docpdag=0) {
    if (docpdag) x <- cpdag(x) # here cpdag
    y <- graph2adj(as.graphNEL(x))
  },docpdag=docpdag)
  bnbs <- do.call("abind",list(bnbs,along=3))
  bnbs <- apply(bnbs,c(1,2),sum)/n
  bnorg <- bnorg*bnbs
  nodecol <- list()
  for (i in 1:nrow(bnorg)) {
    nodecol[[rownames(bnorg)[i]]] <- rgb(0.7,0.7,0.7)
    if (rownames(bnorg)[i] %in% useVars) nodecol[[rownames(bnorg)[i]]] <- rgb(1,0,0)
  }
  nodecol[[response]] <- rgb(1,1,1)
  edgelabels <- character(length(bnbs))
  for (i in 1:length(bnbs)) {
    tmp <- binom.test(round(t(bnbs)[i]*n),n)
    edgelabels[i] <- paste0(round(t(bnbs)[i]*n),"\n",
                            "[",round(tmp$conf.int[1],3),";",round(tmp$conf.int[2],3),"]")
  }
  bnbscut <- bnbs
  bnbscut[bnbs<m] <- 0
  bnbscutdnf <- adj2dnf(bnbscut)
  plotDnf(adj2dnf(bnbscut),nodeshape="box",width=5,nodecol=nodecol,
          main="boostrapped data cut off by 0.5 with support and CI",
          edgelabel=c(rep("",length(bnbscutdnf)-length(grep("=",bnbscutdnf))),
                      edgelabels[which(as.vector(t(bnbscut))!=0)]),showall=FALSE)
  cat(".")
}
```

```{r,echo=TRUE}
stopCluster(cl)
sessionInfo()
end <- Sys.time()
print("Runtime:")
print(end-start)
```



















